---
title: "Palette gallery"
description: "Every color option registered in scChromatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Palette gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(scChromatic)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

This gallery is generated from the package's offline registry, so it stays in
sync with the colors returned by `sc_palette()`. Hover over a qualitative
swatch to see its hexadecimal value.

```{r summary, results='asis'}
ids <- sc_palette_names()
info <- do.call(rbind, lapply(ids, sc_palette_info))
counts <- table(info$palette_type)
cat(
  '<div class="palette-summary">',
  sprintf("<div><strong>%d</strong><span>registered palettes</span></div>", nrow(info)),
  sprintf(
    "<div><strong>%d</strong><span>qualitative</span></div>",
    unname(counts[["qualitative"]])
  ),
  sprintf(
    "<div><strong>%d</strong><span>sequential</span></div>",
    unname(counts[["sequential"]])
  ),
  sprintf(
    "<div><strong>%d</strong><span>diverging</span></div>",
    unname(counts[["diverging"]])
  ),
  "</div>",
  sep = ""
)
```

```{r helpers, include=FALSE}
palette_card <- function(id) {
  metadata <- sc_palette_info(id)
  type <- metadata$palette_type[[1L]]
  if (startsWith(id, "scico_") &&
      !requireNamespace("scico", quietly = TRUE)) {
    return(paste0(
      '<article class="palette-card">',
      '<div class="palette-id"><strong class="palette-heading"><code>',
      id,
      "</code></strong>",
      '<span class="palette-badge">', type, "</span></div>",
      '<div class="palette-meta">',
      metadata$source[[1L]], " · ", metadata$status[[1L]],
      "<br>", gsub(",", " · ", metadata$intended_use[[1L]]),
      "</div>",
      '<div class="palette-unavailable">',
      "Install the optional scico package to render this palette.",
      "</div></article>"
    ))
  }
  n <- if (type == "qualitative") metadata$max_n[[1L]] else 64L
  colors <- sc_palette(
    id,
    n = n,
    selection = "source",
    extend = "generate"
  )
  preview <- if (type == "qualitative") {
    swatches <- paste0(
      '<span class="palette-swatch" style="background:',
      colors,
      '" title="',
      seq_along(colors),
      ": ",
      colors,
      '"></span>',
      collapse = ""
    )
    paste0('<div class="palette-swatches">', swatches, "</div>")
  } else {
    paste0(
      '<div class="palette-gradient" style="background:linear-gradient(90deg,',
      paste(colors, collapse = ","),
      ')"></div>'
    )
  }
  paste0(
    '<article class="palette-card">',
    '<div class="palette-id"><strong class="palette-heading"><code>',
    id,
    "</code></strong>",
    '<span class="palette-badge">', type, "</span></div>",
    '<div class="palette-meta">',
    metadata$source[[1L]], " · ",
    if (type == "qualitative") paste0(metadata$max_n[[1L]], " colors · ") else "",
    metadata$status[[1L]],
    "<br>", gsub(",", " · ", metadata$intended_use[[1L]]),
    "</div>",
    preview,
    "</article>"
  )
}

palette_section <- function(title, types) {
  section <- info$palette_id[info$palette_type %in% types]
  cat(
    "<h2>", title, "</h2>",
    '<div class="palette-grid">',
    paste(vapply(section, palette_card, character(1)), collapse = ""),
    "</div>",
    sep = ""
  )
}
```

```{r qualitative, results='asis'}
palette_section("Qualitative palettes", "qualitative")
```

Qualitative palettes are fixed sequences. scChromatic does not interpolate
them by default; request `extend = "generate"` only when more colors are
needed.

```{r sequential, results='asis'}
palette_section("Sequential palettes", "sequential")
```

```{r diverging, results='asis'}
palette_section("Diverging palettes", "diverging")
```

For signed values, pass the data midpoint to the scale so rescaling is
centered correctly:

```{r signed-example, echo=TRUE}
scale_color_sc_c("archr_coolwarm", midpoint = 0)
```

High-cardinality palettes are perceptually optimized or compatibility
collections, not guarantees of universal color-vision-deficiency safety. Use
`sc_palette_audit()` and combine color with labels, shapes, facets, or other
encodings when distinctions matter.
