---
title: "Palette accessibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Palette accessibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
```

## Normal and simulated CVD views

```{r}
audit <- sc_palette_audit(
  "okabe_ito",
  cvd = c("none", "deutan", "protan", "tritan")
)
audit

views <- sc_palette_plot(
  "okabe_ito",
  view = "points",
  cvd = c("none", "deutan", "protan", "tritan")
)
views$none
views$deutan
```

The audit reports minimum and median CIE2000 distances in every requested
vision simulation plus contrast against the chosen background.

## Light and dark backgrounds

```{r}
sc_palette_audit("tol_muted", background = "#FFFFFF", cvd = "none")
sc_palette_audit("tol_muted", background = "#1A1A1A", cvd = "none")
sc_palette_recommend(
  8, use = "cell_identity", geometry = "point", background = "dark"
)
```

## Point versus fill geometry

Small dense points are harder to distinguish than broad filled regions.
Registry recommendations therefore include geometry and background metadata;
contrast values are diagnostics rather than automatic WCAG pass/fail claims for
plot marks.

```{r}
sc_palette_plot("chromatic", n = 12, view = "points", cvd = "none")
sc_palette_plot("chromatic", n = 12, view = "swatch", cvd = "none")
```

## High-cardinality limitations

No 30- or 40-color sequence is universally color-blind safe. At high
cardinality, combine color with direct labels, facets, shapes, line types,
spatial separation, or interactive lookup. Use hierarchy maps when subtype
similarity should be explicit.

```{r}
sc_hierarchy_map(
  parent = c("Lymphoid", "Lymphoid", "Lymphoid", "Myeloid", "Myeloid"),
  child = c("B", "T", "NK", "Mono", "DC")
)
```
