---
title: "scChromatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scChromatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
library(ggplot2)
data(sc_example)
```

## Palette discovery

The registry separates palette type, intended use, provenance, capacity, and
status.

```{r}
head(sc_palette_names(type = "qualitative"))
sc_palette_info("chromatic")
sc_palette_recommend(12, use = "cell_identity")
```

Raw colors are available as vectors, while `sc_pal()` returns a
scales-compatible closure.

```{r}
sc_palette("chromatic", 6)
sc_pal("chromatic")(6)
```

## Example single-cell dataset

`sc_example` is deterministic and synthetic. Its columns represent the color
semantics commonly encountered in single-cell figures.

```{r}
dim(sc_example)
head(sc_example)
```

## Persistent cell identities

```{r}
cell_map <- sc_color_map(sc_example$cell_type, palette = "chromatic")

ggplot(sc_example, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.5) +
  scale_color_sc_map(cell_map)
```

The same map is reused after filtering, so the three retained identities keep
their original colors.

```{r}
stimulated <- subset(
  sc_example,
  condition == "Stimulated" & cell_type %in% c("CD4 T", "NK", "Monocyte")
)

ggplot(stimulated, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.7) +
  scale_color_sc_map(cell_map)
```

## Choosing a ggplot2 scale

Choose a scale from both the mapped data and the ggplot2 aesthetic:

| Mapped data and assignment strategy | `color` or `colour` aesthetic | `fill` aesthetic |
|:--|:--|:--|
| Categorical, assigned for one plot | `scale_color_sc_d()` | `scale_fill_sc_d()` |
| Categorical, locked in an `sc_color_map` | `scale_color_sc_map()` | `scale_fill_sc_map()` |
| Numeric, shown as a gradient | `scale_color_sc_c()` | `scale_fill_sc_c()` |

The `color` aesthetic controls points, lines, text, and geometry outlines;
`fill` controls the interiors of violins, bars, tiles, polygons, and other
filled geometries. The `scale_colour_*()` functions are exact aliases of the
corresponding `scale_color_*()` functions. Match the scale to the aesthetic in
`aes()`: mapping `fill = sample` requires a `scale_fill_*()` scale, while
mapping `color = sample` requires a `scale_color_*()` scale.

Use the `_sc_map()` family when categories must keep the same colors across
figures, subsets, or reordered factors. A single map can drive either
aesthetic.

```{r}
sample_map <- sc_color_map(sc_example$sample, palette = "chromatic")

ggplot(sc_example, aes(UMAP1, UMAP2, color = sample)) +
  geom_point(size = 0.5) +
  scale_color_sc_map(sample_map)

ggplot(sc_example, aes(sample, n_counts, fill = sample)) +
  geom_violin() +
  scale_y_log10() +
  scale_fill_sc_map(sample_map) +
  labs(x = NULL, y = "Library size")
```

Use `_sc_d()` for a one-off categorical plot where a persistent assignment is
not needed. Its assignments are computed from the discrete levels supplied to
that plot.

```{r}
ggplot(sc_example, aes(UMAP1, UMAP2, color = condition)) +
  geom_point(size = 0.5) +
  scale_color_sc_d("chromatic")
```

Use `_sc_c()` only for numeric values. Sequential, diverging, or cyclic
palettes are valid for continuous scales; qualitative palettes are not. If a
category is stored as numbers, convert it to a factor before using `_sc_d()` or
constructing an `sc_color_map`.

These rules do not depend on the source object. For a Seurat object,
`object[[]]` supplies its metadata data frame and `object$sample` supplies the
sample vector used to construct a map. A plot returned by another package,
including pixelatorR, must still use the scale matching the aesthetic mapped by
that plot. If both `color` and `fill` are mapped, add one scale for each
aesthetic; otherwise, use only the scale that matches `aes()`.

## Continuous expression

```{r}
ggplot(sc_example, aes(UMAP1, UMAP2, color = MS4A1)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("viridis")
```

## Signed scores centered at zero

`midpoint = 0` changes data rescaling, not merely the middle displayed color.

```{r}
ggplot(sc_example, aes(UMAP1, UMAP2, color = signed_score)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("chromatic_balance", midpoint = 0)
```

## Parent lineage and child subtype colors

```{r}
lineage_map <- sc_hierarchy_map(sc_example$lineage, sc_example$cell_type)

ggplot(sc_example, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.5) +
  scale_color_sc_map(lineage_map)
```

## Pseudotime and QC

```{r}
ggplot(sc_example, aes(UMAP1, UMAP2, color = pseudotime)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("cividis")

ggplot(sc_example, aes(UMAP1, UMAP2, color = percent_mito)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("viridis")
```
