---
title: "scChromatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scChromatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
library(ggplot2)
set.seed(1)
```

## Palette discovery

The registry separates palette type, intended use, provenance, capacity, and
status.

```{r}
head(sc_palette_names(type = "qualitative"))
sc_palette_info("archr_stallion")
sc_palette_recommend(12, use = "cell_identity")
```

Raw colors are available as vectors, while `sc_pal()` returns a
scales-compatible closure.

```{r}
sc_palette("tol_muted", 6)
sc_pal("chromatic")(6)
```

## Synthetic single-cell embedding

```{r}
n <- 800
cell_types <- c("B", "T", "NK", "Mono")
umap <- data.frame(
  UMAP1 = rnorm(n) + rep(c(-2, 2, -1, 1), each = n / 4),
  UMAP2 = rnorm(n) + rep(c(1, 1, -2, -1), each = n / 4),
  cell_type = rep(cell_types, each = n / 4)
)

ggplot(umap, aes(UMAP1, UMAP2, color = cell_type)) +
  geom_point(size = 0.5) +
  scale_color_sc_d("chromatic")
```

## ArchR palettes

ArchR values are frozen from commit
`6feec354ad6c8052ddbc4626a2ca2d858ed465bf`. Choose source order for literal
upstream order or priority order for nested subsets.

```{r}
sc_palette("archr_stallion", 8, selection = "source")
pal_archr("stallion", selection = "priority")(8)
```

## Continuous expression

```{r}
umap$expression <- pmax(0, umap$UMAP1 + rnorm(n))
ggplot(umap, aes(UMAP1, UMAP2, color = expression)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("viridis")
```

## Signed scores centered at zero

`midpoint = 0` changes data rescaling, not merely the middle displayed color.

```{r}
umap$score <- umap$UMAP1 * umap$UMAP2
ggplot(umap, aes(UMAP1, UMAP2, color = score)) +
  geom_point(size = 0.5) +
  scale_color_sc_c("archr_coolwarm", midpoint = 0)
```
