---
title: "Stable color maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stable color maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
library(ggplot2)
```

## Define the mapping from the full annotation

```{r}
labels_full <- c("B", "T", "NK", "Mono")
map <- sc_color_map(labels_full, palette = "chromatic")
as_named_colors(map)
```

Subsets select names from the same object; they never rerun positional palette
assignment.

```{r}
labels_subset <- c("T", "NK")
subset_colors <- as_named_colors(map)[labels_subset]
stopifnot(identical(
  subset_colors,
  as_named_colors(map)[c("T", "NK")]
))
```

Factor reordering also leaves the contract intact.

```{r}
reordered <- factor(labels_full, levels = rev(labels_full))
as_named_colors(map)[levels(reordered)]
```

## Align colors for AnnData

AnnData stores categorical colors positionally in
`.uns['<category>_colors']`; that vector must use the same order as the
corresponding categorical levels. Keep the mapping named in R, then select it
in factor-level order immediately before writing it to AnnData.

```{r}
cell_type <- factor(labels_full, levels = c("T", "NK", "B", "Mono"))
ann_data_colors <- as_named_colors(map)[levels(cell_type)]
ann_data_colors
```

`unname(ann_data_colors)` can be written to `.uns['cell_type_colors']` with
zellkonverter, reticulate, or another AnnData interface. If levels are reordered,
repeat the indexed `as_named_colors(map)[levels(cell_type)]` lookup rather than
reusing a positional vector. This keeps the AnnData colors aligned without
adding an AnnData dependency to scChromatic.

CELLXGENE also uses the cyclic `d3_rainbow` scheme as a categorical fallback,
but those assignments change with the category count and order. For that
reason, `d3_rainbow` is a compatibility palette and is intentionally unavailable
to `sc_color_map()`; use a fixed qualitative palette for persistent mappings.

## Add a new annotation

```{r}
updated <- update_sc_color_map(map, c(labels_full, "DC"))
stopifnot(identical(
  as_named_colors(updated)[names(as_named_colors(map))],
  as_named_colors(map)
))
```

## Export and restore

JSON is used when jsonlite is installed. CSV is always a transparent fallback.

```{r}
path <- tempfile(fileext = ".json")
write_sc_color_map(updated, path)
restored <- read_sc_color_map(path)
stopifnot(identical(as_named_colors(restored), as_named_colors(updated)))
```

## Plot and framework interoperability

```{r}
dat <- data.frame(
  x = seq_along(labels_full),
  y = c(1, 3, 2, 4),
  cell_type = labels_full
)
ggplot(dat, aes(x, y, color = cell_type)) +
  geom_point(size = 4) +
  scale_color_sc_map(map)
```

`as_named_colors(map)` is a plain named character vector. It can be passed to
Seurat manual scales, SingleCellExperiment plotting code, ArchR color
arguments, or `ComplexHeatmap::HeatmapAnnotation()` without loading any of
those frameworks in scChromatic.

Frozen ArchR compatibility colors remain available under explicit `archr_*`
palette IDs, but stable mappings do not depend on them.
