---
title: "Stable color maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stable color maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
library(ggplot2)
```

## Define the mapping from the full annotation

```{r}
labels_full <- c("B", "T", "NK", "Mono")
map <- sc_color_map(labels_full, palette = "chromatic")
as_named_colors(map)
```

Subsets select names from the same object; they never rerun positional palette
assignment.

```{r}
labels_subset <- c("T", "NK")
subset_colors <- as_named_colors(map)[labels_subset]
stopifnot(identical(
  subset_colors,
  as_named_colors(map)[c("T", "NK")]
))
```

Factor reordering also leaves the contract intact.

```{r}
reordered <- factor(labels_full, levels = rev(labels_full))
as_named_colors(map)[levels(reordered)]
```

## Align colors for AnnData

AnnData stores categorical colors positionally in
`.uns['<category>_colors']`; that vector must use the same order as the
corresponding categorical levels. Keep the mapping named in R, then select it
in factor-level order immediately before writing it to AnnData.

```{r}
cell_type <- factor(labels_full, levels = c("T", "NK", "B", "Mono"))
ann_data_colors <- as_named_colors(map)[levels(cell_type)]
ann_data_colors
```

`unname(ann_data_colors)` can be written to `.uns['cell_type_colors']` with
zellkonverter, reticulate, or another AnnData interface. If levels are reordered,
repeat the indexed `as_named_colors(map)[levels(cell_type)]` lookup rather than
reusing a positional vector. This keeps the AnnData colors aligned without
adding an AnnData dependency to scChromatic.

CELLXGENE also uses the cyclic `d3_rainbow` scheme as a categorical fallback,
but those assignments change with the category count and order. For that
reason, `d3_rainbow` is a compatibility palette and is intentionally unavailable
to `sc_color_map()`; use a fixed qualitative palette for persistent mappings.

## Add a new annotation

```{r}
updated <- update_sc_color_map(map, c(labels_full, "DC"))
stopifnot(identical(
  as_named_colors(updated)[names(as_named_colors(map))],
  as_named_colors(map)
))
```

Hierarchy maps use the same append-only rule when they are extended through
`sc_hierarchy_map(existing = ...)`. New parents and children receive colors,
while established child colors and parent anchors remain unchanged.

```{r}
hierarchy <- sc_hierarchy_map(
  parent = c("Lymphoid", "Lymphoid", "Myeloid"),
  child = c("B", "T", "Mono")
)
hierarchy_updated <- sc_hierarchy_map(
  parent = c("Lymphoid", "Myeloid"),
  child = c("NK", "DC"),
  existing = hierarchy
)
stopifnot(identical(
  as_named_colors(hierarchy_updated)[names(as_named_colors(hierarchy))],
  as_named_colors(hierarchy)
))
```

## Derive context from coordinate neighborhoods

`sc_relationship_map()` accepts a symmetric affinity matrix in `[0, 1]`.
Affinities near one ask related labels to use perceptually closer—but still
distinct—colors; affinities near zero ask unrelated labels to be farther apart.
The optimizer fits normal-vision CIE2000 distance to the supplied or derived
affinity target. A separate penalty uses the worst distance across normal, deutan,
protan, and tritan simulations by default, penalizing shortfalls below that
target in any requested view.

`sc_relationship_from_knn()` can derive the matrix from coordinates. It builds
an exact Euclidean k-nearest-neighbor graph within each sample, symmetrizes
label-to-label neighbor proportions, and combines sample matrices with equal
weight. A sample contributes to a pair only when both labels are present; this
pairwise-complete rule prevents missing labels from being treated as evidence
of separation. A pair never observed together is an error unless
`unobserved = "zero"` explicitly makes that assumption. Use
`aggregate = "median"` instead of the default mean when a median across
contributing samples is more appropriate.

```{r}
data(sc_example)
affinity <- sc_relationship_from_knn(
  as.matrix(sc_example[c("UMAP1", "UMAP2")]),
  labels = sc_example$cell_type,
  sample = sc_example$sample,
  k = 15,
  aggregate = "mean"
)

canonical <- sc_color_map(sc_example$cell_type)
relationship_map <- sc_relationship_map(
  affinity,
  canonical = canonical,
  locked = "B cell",
  stability_budget = 1,
  seed = 2026
)
as_named_colors(relationship_map)
```

The bundled data expose UMAP-like coordinates, so they are used here only to
demonstrate the interface. UMAP does not preserve all quantitative distances.
For scientific neighborhood context, prefer appropriately scaled PCA, a
model-derived latent space, or spatial coordinates, and report the preprocessing
and `k`. Equal sample weights prevent a larger sample from dominating, but do
not by themselves correct batch effects.
Neighborhood affinities also depend on label prevalence, local density,
coordinate scaling, and `k`; they should not be interpreted as direct evidence
of biological similarity. Stored provenance retains fingerprints and aggregate
sample summaries, not raw coordinates or sample identifiers.

Hard-locked labels never change. The stability budget is a separate maximum on
how many other canonical assignments may change; new labels do not consume it.
The output is an ordinary derived `sc_color_map`, so the same scales and
export functions apply. Its `context`, `history`, and `metadata` retain the
method version, seed, requested CVD modes, objective, normalized effective
upper-triangle affinities, canonical before/after colors, and changes. These
records describe construction and remain immutable if the map is subset. To
add labels or recalculate metrics for a subset, rerun `sc_relationship_map()`
with the corresponding matrix; generic map extension is rejected because it
would leave the stored relationship input stale.

## Add shape or pattern redundancy

Color need not carry the full annotation alone. `sc_redundant_encoding()` finds
pairs below a chosen worst-vision CIE2000 distance across normal and requested
CVD simulations, then assigns different auxiliary encodings to those pairs.
The original colors are never changed.

```{r}
encoding <- sc_redundant_encoding(
  relationship_map,
  channel = "shape",
  min_cie2000 = 8
)
shape_values <- setNames(encoding$shape, encoding$label)

ggplot(sc_example, aes(UMAP1, UMAP2, color = cell_type, shape = cell_type)) +
  geom_point(size = 1) +
  scale_color_sc_map(relationship_map) +
  scale_shape_manual(values = shape_values)
```

For geometries that can render texture, use `channel = "pattern"` or
`channel = "both"`. The returned `pattern` names are suitable for compatible
pattern layers, while `texture_group` is a generic group index for texture
systems such as scatterHatch. These integrations remain optional; scChromatic
does not require a pattern-rendering package.
The allocation is one-shot: create it for the full label universe and reuse or
subset the same table across figures. Rerunning after labels are added or
removed can change DSATUR priorities. RDS preserves the table's method and
conflict attributes; ordinary CSV export does not.

## Export and restore

JSON follows the installed versioned color-map schema. CSV keeps readable
label/color rows and embeds the same complete JSON payload in one metadata row.
Both formats preserve provenance and hierarchy, focus, alias, lock, history,
context, and seed metadata. Legacy unversioned JSON and two-column CSV files are
migrated when read; unknown future schema versions are rejected.

```{r}
path <- tempfile(fileext = ".json")
write_sc_color_map(updated, path)
restored <- read_sc_color_map(path)
stopifnot(identical(as_named_colors(restored), as_named_colors(updated)))
```

## Plot and framework interoperability

```{r}
dat <- data.frame(
  x = seq_along(labels_full),
  y = c(1, 3, 2, 4),
  cell_type = labels_full
)
ggplot(dat, aes(x, y, color = cell_type)) +
  geom_point(size = 4) +
  scale_color_sc_map(map)
```

`as_named_colors(map)` is a plain named character vector. It can be passed to
Seurat manual scales, SingleCellExperiment plotting code, ArchR color
arguments, or `ComplexHeatmap::HeatmapAnnotation()` without loading any of
those frameworks in scChromatic.
