---
title: "Stable color maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stable color maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(scChromatic)
library(ggplot2)
```

## Define the mapping from the full annotation

```{r}
labels_full <- c("B", "T", "NK", "Mono")
map <- sc_color_map(labels_full, palette = "chromatic")
as_named_colors(map)
```

Subsets select names from the same object; they never rerun positional palette
assignment.

```{r}
labels_subset <- c("T", "NK")
subset_colors <- as_named_colors(map)[labels_subset]
stopifnot(identical(
  subset_colors,
  as_named_colors(map)[c("T", "NK")]
))
```

Factor reordering also leaves the contract intact.

```{r}
reordered <- factor(labels_full, levels = rev(labels_full))
as_named_colors(map)[levels(reordered)]
```

## Align colors for AnnData

AnnData stores categorical colors positionally in
`.uns['<category>_colors']`; that vector must use the same order as the
corresponding categorical levels. Keep the mapping named in R, then select it
in factor-level order immediately before writing it to AnnData.

```{r}
cell_type <- factor(labels_full, levels = c("T", "NK", "B", "Mono"))
ann_data_colors <- as_named_colors(map)[levels(cell_type)]
ann_data_colors
```

`unname(ann_data_colors)` can be written to `.uns['cell_type_colors']` with
zellkonverter, reticulate, or another AnnData interface. If levels are reordered,
repeat the indexed `as_named_colors(map)[levels(cell_type)]` lookup rather than
reusing a positional vector. This keeps the AnnData colors aligned without
adding an AnnData dependency to scChromatic.

CELLXGENE also uses the cyclic `d3_rainbow` scheme as a categorical fallback,
but those assignments change with the category count and order. For that
reason, `d3_rainbow` is a compatibility palette and is intentionally unavailable
to `sc_color_map()`; use a fixed qualitative palette for persistent mappings.

## Add a new annotation

```{r}
updated <- update_sc_color_map(map, c(labels_full, "DC"))
stopifnot(identical(
  as_named_colors(updated)[names(as_named_colors(map))],
  as_named_colors(map)
))
```

Hierarchy maps use the same append-only rule when they are extended through
`sc_hierarchy_map(existing = ...)`. New parents and children receive colors,
while established child colors and parent anchors remain unchanged.

```{r}
hierarchy <- sc_hierarchy_map(
  parent = c("Lymphoid", "Lymphoid", "Myeloid"),
  child = c("B", "T", "Mono")
)
hierarchy_updated <- sc_hierarchy_map(
  parent = c("Lymphoid", "Myeloid"),
  child = c("NK", "DC"),
  existing = hierarchy
)
stopifnot(identical(
  as_named_colors(hierarchy_updated)[names(as_named_colors(hierarchy))],
  as_named_colors(hierarchy)
))
```

## Export and restore

JSON follows the installed versioned color-map schema. CSV keeps readable
label/color rows and embeds the same complete JSON payload in one metadata row.
Both formats preserve provenance and hierarchy, focus, alias, lock, history,
context, and seed metadata. Legacy unversioned JSON and two-column CSV files are
migrated when read; unknown future schema versions are rejected.

```{r}
path <- tempfile(fileext = ".json")
write_sc_color_map(updated, path)
restored <- read_sc_color_map(path)
stopifnot(identical(as_named_colors(restored), as_named_colors(updated)))
```

## Plot and framework interoperability

```{r}
dat <- data.frame(
  x = seq_along(labels_full),
  y = c(1, 3, 2, 4),
  cell_type = labels_full
)
ggplot(dat, aes(x, y, color = cell_type)) +
  geom_point(size = 4) +
  scale_color_sc_map(map)
```

`as_named_colors(map)` is a plain named character vector. It can be passed to
Seurat manual scales, SingleCellExperiment plotting code, ArchR color
arguments, or `ComplexHeatmap::HeatmapAnnotation()` without loading any of
those frameworks in scChromatic.
